An Accurate Model for SARS-CoV-2 pooled RT-PCR Tests Errors

### 2021-03-16 comment.tex
Better even more gooder

%% Before we proceed, it is important to distinguish three types of
%% false-negative events when performing pooling. For convenience, we
%% follow a single \emph{infected} individual, henceforth referred to as
%% "Donald". A \emph{single test's} false-negative is the event of a
%% negative result upon testing Donald separately, i.e., in an RT-PCR
%% test without pooling. We denote the test sensitivity $\Se$, so the
%% probability of a single test false-negative is $1-\Se$. A
%% \emph{pooled} false-negative occurs when a pooled test containing
%% Donald's sample (and other samples) yields a negative result, i.e.,
%% the pooling fails to detect at least one positive result. Lastly, a
%% \emph{scheme} false-negative occurs when an entire pooling scheme
%% fails to identify Donald as infected. Our goal is to calculate
%% Dorfman's scheme false-negative rate. Rephrasing, we wish to answer
%% the following question: what is the probability of not identifying
%% Donald as infected under a Dorfman pooling scheme?

@@

Formally, we consider a pool containing $N$ individuals
$\{1,\dots,N\}$. We denote the true infection states $\theta \in
\{0,1\}^N$, so individual $i$ is infected iff $\theta_i=1$. The RT-PCR
test's sensitivity (true-negative rate) is denoted $\sens$, and the
test's specificity (true-negative rate) is denoted $\spec$. Test results
(data) are denoted $\data$, where $\data=1$ iff the test returned a
positive result.

@@

\begin{align}\label{eq:common}
\Pr(\data=1|\theta) = \left \{
\begin{array}{lr}
        \sens & \exists i\ \text{such that } \theta_i = 1\\
        1-\spec & \text{otherwise}
\end{array} 
\end{align}
We refute the assumptions underlying \eqref{eq:common} in
Section~\ref{subsec:ours} and suggest more plausible ones in
Section~\ref{subsec:ours}.

@@

Erroneous detection of SARS-CoV-2 RNA (a false-positive) in pooled
RT-PCR testing can also occur. A common assumption \cite{Simplistic1,
  Simplistic2, Kim, OptimalDorfmanPool} is that the false-positive
probability does not depend on the number of negative samples in a
pool. We incorporate this assumption in our likelihood model with a
small modification. We assume that an erroneous amplification can
occur in \emph{any} pool. Specifically, it is possible that correct
amplifications fail and an erroneous one occurs simultaneously. This
assumption is relatively specific for the current application of
screening for SARS-CoV-2 via RT-PCR.  For example, cross-reactivity
with other coronaviruses would have violated this assumption, but it
was ruled out in \cite{KitComparison}.

@@

      \node[shape=rectangle,draw=black] (G) at (4,7)
           {$\Pr(\data=0|\theta)=\fn^2(1-\fp)$};
    \end{scope}

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=red,very thick}]
      \path [->] (A) edge node {$\fn$} (4.5,6.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (E) edge node {$\fn$} (4.5,6.6);
      \path [->] (F) edge node {$1-\fp$} (5.5,6.6); 
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of the likelihood model of a pooled test
    result. A pool contains individuals $\{1,2,3,4\}$ with state
    $\theta=(1,0,0,1)$ (i.e. individuals 1 and 4 are infected). A
    negative pooled test implies that three detection paths failed. A
    false negative occurred for $1$ and $4$, each with probability
    $\fn$. Additionally, no erroneous detection of SARS-CoV-2
    occurred, with probability $1-\fp$. Individuals $2$ and $3$ are
    not infected and do not contribute to the probability of the
    pooled test result. A positive pooled test arises if any one of
    the above mentioned paths results in a
    detection.}\label{fig:likelihood}
\end{figure}

@@

The probability of a negative pooled test result is presented in
\eqref{eq:likelihood a} (along with its complement
\eqref{eq:likelihood b}), and explained below.
\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= (1-\fp)\prod_{i=1}^N
    \fn^{\theta_i}, \label{eq:likelihood a}\\
    %
    %
    %
    \Pr(\data=1 | \theta) &= 1 - (1-\fp)\prod_{i=1}^N \fn^{\theta_i}. \label{eq:likelihood b}
  %\end{split}
\end{align}
\end{subequations}
A negative pooled test result occurs when all detection paths (both
correct and erroneous) fail. The probability of no false-detection
accounts for the $1-\fp$ term. The probability of no correct detection
is $\fn$ per infected individual. The probability that all such paths
fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}. The probability of a positive result,
presented in \eqref{eq:likelihood b}, is simply the complement.
Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
recalling that $\data \in \{0,1\}$ yields:

\begin{align}\label{eq:combined likelihood}
  \begin{split}
    \Pr(\data| \theta) &= \left [(1-\fp)\prod_{i=1}^N
      \fn^{\theta_i} \right ]^{1-\data} \left [1 - (1-\fp)\prod_{i=1}^N \fn^{\theta_i} \right ]^{\data}.
  \end{split}
\end{align}

\section{Application: Calculation of Dorfman's scheme false-negative rate}\label{subsec:detailed}
As an application of our model, we present the differnece in scheme
false-negative rates (...) Denote the prevalence of infection in the
(tested) population $q$. As before, $\Se$ is the test's sensitivity,
so $1-\Se$ is the single RT-PCR test's false-negative rate. We also
denote $\Sp$ the test specificity --- the true-negative
probability. This encompasses events of an erroneous RNA detection
(via any possible pathway), which may cause a false-positive. By our
assumptions, a pool containing Donald's sample and $N-1$ other samples
will yield a \emph{negative} result if all of the following occur:
\begin{itemize}
\item No erroneous detection occurred (i.e. no false-positive). This
  happens with probability $\Sp$.
\item The detection process fails for Donald's sample. A
  false-negative occurs for Donald, with probability $1-\Se$.
\item No detection for any of the other $N-1$ samples. For a single
  sample, the probability of being detected is the prevalence of
  SARS-CoV-2 in the tested population $q$, multiplied by the
  sensitivity $\Se$, hence the probability of detection is $q\Se$. For
  $N-1$ such samples, the probability of \emph{not} being identified
  is $(1-q\Se)^{N-1}$.
\end{itemize}
The pooled false-negative probability for Donald is simply the product of the terms above. Hence:

\begin{align}
    \begin{split}
        \Pr(\data=1|\theta_1=1) &= 1 - \Pr(\data=0|\theta_1=1) \\
        &= 1 - \Sp(1-\Se)(1-q\Se)^{N-1}.
    \end{split}
\end{align}

If the pooled test yields a positive result, Donald is tested
separately. We assume such a simple procedure poses no risk of
introducing contaminant RNA. Therefore, the separate test yields a
positive result with probability $\Se$.

We calculate the probability that Donald is mistakenly identified as
not infected --- the scheme's false-negative rate --- denoted $\mi$
below. To \emph{correctly} identify Donald as infected, both pooled
and separate tests have to yield a positive result. Thus, the scheme's
false-negative rate $\mi$ is the complement of the product of the two
previous terms:
\begin{align}\label{eq:sfn}
    \begin{split}
        \mi :&= 1-\Pr(\text{correctly identify Donald as infected})\\
        %
        &= 1 - \Se\left [1 - \Sp(1-\Se)(1-q\Se))^{N-1}\right].
    \end{split}
\end{align}

@@

For the approximation commonly used in the literature estimates the
scheme false-negative rate as $1-\Se^2$ (Section
\ref{subsec:assumptions} and \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}). A short calculation shows that this
approximation implies the percentage increase in scheme false-negative
rate is just $100\cdot \Se$.

@@

\bibliographystyle{ama}
\bibliography{refs}

>>>>

@@

Formally, we consider a pool containing $N$ individuals
$\{1,\dots,N\}$. We denote the true infection states $\theta \in
\{0,1\}^N$, so individual $i$ is infected iff $\theta_i=1$. The RT-PCR
test's sensitivity (true-negative rate) is denoted $\sens$, and the
test's specificity (true-negative rate) is denoted $\spec$. Test results
(data) are denoted $\data$, where $\data=0$ iff the test returned a
positive result.

@@

\begin{align}\label{eq:common}
\Pr(\data=0|\theta) = \left \{
\begin{array}{lr}
        1-\sens & \exists i\ \text{such that } \theta_i = 1\\
        \spec & \text{otherwise}
\end{array} 
\end{align}
We refute the assumptions underlying \eqref{eq:common} in
Section~\ref{subsec:ours} and suggest more plausible ones in
Section~\ref{subsec:ours}.

@@

\subsection{Our model}
We assume two pathways for a positive pooled test result: Viral RNA
from an infected individual is correctly detected; or, some erroneous
detection occurs (e.g. contaminant viral RNA is introduced). We assume
a homogeneous and disconnected population (each individual is infected
independently and with equal probability. Erroneous detection of
SARS-CoV-2 RNA (a false-positive) in pooled RT-PCR testing can also
occur. A common assumption \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool} is that the false-positive probability does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our likelihood model with a small modification. We
assume that an erroneous amplification can occur in \emph{any}
pool. Specifically, it is possible that correct amplifications fail
and an erroneous one occurs simultaneously. This assumption is
relatively specific for the current application of screening for
SARS-CoV-2 via RT-PCR.  For example, cross-reactivity with other
coronaviruses would have violated this assumption, but it was ruled
out in \cite{KitComparison}.

@@

      \node[shape=rectangle,draw=black] (G) at (4,7)
           {$\Pr(\data=0|\theta)=(1-\sens)^2\spec$};
    \end{scope}

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=red,very thick}]
      \path [->] (A) edge node {$1-\sens$} (4.5,6.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (E) edge node {$1-\sens$} (4.5,6.6);
      \path [->] (F) edge node {$\spec$} (5.5,6.6); 
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of the likelihood model of a pooled test
    result. A pool contains individuals $\{1,2,3,4\}$ with state
    $\theta=(1,0,0,1)$ (i.e. individuals 1 and 4 are infected). A
    negative pooled test implies that three detection paths failed. A
    false negative occurred for $1$ and $4$, each with probability
    $(1-\sens)$. Additionally, no erroneous detection of SARS-CoV-2
    occurred, with probability $\spec$. Individuals $2$ and $3$ are
    not infected and do not contribute to the probability of the
    pooled test result. A positive pooled test arises if any one of
    the above mentioned paths results in a
    detection.}\label{fig:likelihood}
\end{figure}

@@

A negative pooled test result occurs when all detection paths (both
correct and erroneous) fail. The probability of no false-detection
accounts for the $\spec$ term. The probability of no correct detection
is $(1-\sens)$ per infected individual. The probability that all such
paths fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}.%%  The probability of a positive result,
%% presented in \eqref{eq:likelihood b}, is simply the complement.
%% The probability of a negative pooled test result is presented in
%% \eqref{eq:likelihood a} (along with its complement
%% \eqref{eq:likelihood b}), and explained below.
%\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum\theta_i}, \label{eq:likelihood a}
    %
    %
    %
    %% \Pr(\data=1 | \theta) &= 1 - \spec\prod_{i=1}^N
    %% (1-\sens)^{\theta_i} = 1 - \spec(1-\Se)^{\sum\theta_i} . \label{eq:likelihood b}
  %\end{split}
\end{align}
%\end{subequations}
%% Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
%% recalling that $\data \in \{0,1\}$ yields:

%% \begin{align}\label{eq:combined likelihood}
%%   \begin{split}
%%     \Pr(\data| \theta) &= \left [\spec\prod_{i=1}^N
%%       (1-\sens)^{\theta_i} \right ]^{1-\data} \left [1 - \spec\prod_{i=1}^N (1-\sens)^{\theta_i} \right ]^{\data}.
%%   \end{split}
%% \end{align}

\section{Application: Calculation of Dorfman's scheme false-negative rate}\label{subsec:detailed}
As an application of our model, we present the differnece in
calculated false-negative rate for a single \emph{infected}
individual, henceforth referred to as "Donald".  We distinguish three
types of false-negative events when performing pooling. A \emph{single
  test's} false-negative is the event of a negative result upon
testing Donald separately, i.e., in an RT-PCR test without pooling. A
\emph{pooled} false-negative occurs when a pooled test containing
Donald's sample (and other samples) yields a negative result, i.e.,
the pooling fails to detect at least one positive result. Lastly, a
\emph{scheme} false-negative occurs when an entire pooling scheme
fails to identify Donald as infected. Our goal is to calculate
Dorfman's scheme false-negative rate. Rephrasing, we wish to answer
the following question: what is the probability of not identifying
Donald as infected under a Dorfman pooling scheme?

Denote the prevalence of infection in the (tested) population $q$. We
give Donald the respect he deserves and assign him as individual $1$,
so that $\theta_1=1$. Then:

\begin{align}
  \begin{split}
    \Pr(\data=0|\theta_1=1) &= \sum_{\theta_2,\dots,\theta_N} \Pr(\theta') \Pr(\data=0|\theta_1=1,\theta_2,\dots,\theta_N) \\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\Pr(\sum_{i=2}^N\theta_i=k)\Pr(\data=0|\theta_1=1,\sum_{i=2}^N\theta_i = k)\\
    %
    %
    %
    &= \sum_{k=0}^{N-1}\binom{N-1}{k}q^k(1-q)^{N-1-k} \Sp(1-\Se)^{1+k}\\
    %
    %
    %
    &= \Sp(1-\Se) \sum_{k=0}^{N-1}\binom{N-1}{k}\left(q(1-\Se)\right)^k(1-q)^{N-1-k}\\
    %
    %
    %
    &= \Sp(1-\Se)(1-q\Se)^{N-1}, \text{ by Newton's Binomial Theorem.} \\
  \end{split}
\end{align}

If the pooled test yields a positive result, Donald is tested
separately. Taking a conservative stand, we assume such a simple
procedure poses no risk of introducing contaminant RNA. Therefore, the
separate test yields a positive result with probability $\Se$.

We calculate the probability that Donald is mistakenly identified as
not infected --- the scheme's false-negative rate --- denoted $\mi$
below. To \emph{correctly} identify Donald as infected, both pooled
and separate tests have to yield a positive result. Thus, the scheme's
false-negative rate $\mi$ is the complement of the product of the two
previous terms:
\begin{align}\label{eq:sfn}
    \begin{split}
        \mi :&= 1 - \Se\Pr(\data=1|\theta=1)\\
        %
        &= 1 - \Se\left [1 - \Sp(1-\Se)(1-q\Se))^{N-1}\right].
    \end{split}
\end{align}

@@

The approximation commonly used in the literature estimates the scheme
false-negative rate as $1-\Se^2$ (Section \ref{subsec:assumptions} and
\cite{Simplistic1, Simplistic2, Kim, OptimalDorfmanPool}). A short
calculation shows that this approximation implies the percentage
increase in scheme false-negative rate is just $100\cdot \Se$.

@@

\bibliographystyle{amsplain}
%\bibliographystyle{ama}
\bibliography{refs}
